---
output:
  pdf_document: default
  html_document: default
---
# Multivariate differentiation {#multivariate_differentiation}

In the previous chapter we considered functions such as
$f(x)=x^2+\sin(x)$ in which we take a single variable $x$ and function
$f$ gives us a single value in return.  In this chapter we will
consider a generalization of this concept in which a function can take
more than one variable as input.  Consider the following R idiom:


```{r}
f <- function(x,y){x+y^2}
f(5,3)  # should return 5+3^2=5+9=14
```

See again how we verify our work before moving on.  But before we
start considering calculus on such functions we are going to spend
some time looking at

```{r}
f(1:20,3)    # should return 1:20 + 3^2 = 9,10,11,...,29
f(3, 1:4)    # should return 3 + (1:4)^2 = 4,7,12,19
f(1:10,1:10) # should return 1+1^2,2+2^2,...,10+10^2
```

And here we note that this construction can be as complicated as we
like; functions can have any sensible name and take any number of
arguments, which themselves may be complex R expressions.  For
example:

```{r}
robins_function <- function(x,y,z){cos(x+y) - x^2 + z^x}
robins_function(x=1:3, y=c(2.2,5.3,0.1), z=c(0.4,sin(1),1/f(0.4,pi)))
```

## Multivariate calculus: partial differentials

If we are considering a function such as $f(x,y,z)=x^2 + x*y^3
+x\sin(z)$, then we can differentiate this with respect to either $x$
or $y$ or $z$.  To differentiate with respect to $x$, we treat $y$ and
$z$ as though they were a constant, and to differentiate with respect
to $y$ we treat $x$ and $z$ as though constant, and so on.  These are
usually written as follows.

* $f(x,y,z)=x^2 + xy^3
+x\sin(z)\longrightarrow\frac{\partial f}{\partial x}=2x + y^3+\sin(z)$
* $f(x,y,z)=x^2 + xy^3
+x\sin(z)\longrightarrow\frac{\partial f}{\partial y}=3xy^2$
* $f(x,y,z)=x^2 + xy^3
+x\sin(z)\longrightarrow\frac{\partial f}{\partial z}=x\cos(z)$

Make sure you understand this before moving on.

In multivariate calculus one important identity is the following:

\[
f\left(x+\delta x,y+\delta y\right)=
f\left(x,y\right)
+ \delta x\frac{\partial f}{\partial x}
+ \delta y\frac{\partial f}{\partial y}
+\mbox{small correction}
\]

If function $f()$ takes three arguments we would have:


\[
f\left(x+\delta x,y+\delta y,z+\delta z\right)=
f\left(x,y,z\right)
+ \delta x\frac{\partial f}{\partial x}
+ \delta y\frac{\partial f}{\partial y}
+ \delta z\frac{\partial f}{\partial z}
+\mbox{small correction}
\]

And the general case follows directly.  We will verify that the
formula is correct in the case of $f=f(x,y)$.

\newpage

### Verification

```{r}
f <- function(x,y){       # Function f() takes *two* arguments
   x^2 + x/y              # the formula returned
}                         # remember to close braces
x <- 2                    # just a randomish point
y <- 3                    # (x,y) = c(2,3)
f(x,y)                    # evaluate the function at (x,y)
dx <- 1e-6                # arbitrary small value for delta x
dy <- 1e-6                # arbitrary small value for delta y
f(x+dx,y+dy)              # exact 
f(x,y) + (2*x + 1/y)*dx - (x/y^2)*dy # approximate value
```

the agreement is close.  A more sophisticated example would be

```{r}
f <- function(x){sum(x^2)}
f(1:9)
d <- 1e-4*(9:1)
f(d+ 1:9) - f(1:9)
sum(d *2*(1:9))
```

See if you can figure out what is going on here.  Note that, in R
idiom, `f()` takes a *single* argument which is in this case a
vector of length 9.

## Multivariate root-finding

Given a function $f(x,y)$, sometimes we are interested in finding
roots, that is, values of $x$ and $y$ such that $f(x,y)=0$.  This is a
difficult thing in general when $f$ is a complicated function.  If we
consider a simple example: $f(x,y) = x^2+y^2-4$, then this is easily
solved algebraically and we see that $y=\pm\sqrt{4-x^2}$ which is
recognisable as a circle of radius 2.  However, more complicated
examples cannot be solved algebraically and we need to use numerical
methods.

We will consider $f(x,y) = x^2+y^2+x+4y+\sin(xy)$, a more complicated
function; finding roots of this equation algebraically is more or less
impossible.  So we need visual help.  There are a number of ways of
representing this function, but the first step is to define an array
of values of x and y:

```{r}
xvals <- seq(from = -4, to = +4, len=100)
yvals <- seq(from = -4, to = +4, len=100)
jj <- expand.grid(xvals,yvals)
head(jj)
```

Thus each row of variable `jj` is a pair of x,y values.  To
evaluate our function $f(x,y) = x^2+y^2 + x + 4y + \sin(xy)$
we need idiom like this:

```{r}
f <- function(v){
  x <- v[1]
  y <- v[2]
  return(x^2+y^2 + x + 4*y + sin(x*y))
}
```

and to evaluate it we need to use the `apply()` construction:

```{r}
M <- apply(jj,1,f)     # apply the function f() to jj
M <- matrix(M,100,100) # wrap M up into a matrix 
str(M)                 # str() summarises R objects
```

We can view M using `contour()`:

```{r}
contour(xvals,yvals,M,nlevels=30,xlim=c(-5,5),ylim=c(-5,5),asp=1,axes=FALSE)
axis(1,pos=-5,at=c(-4,-2,0,2,4))
axis(2,pos=-5)
```

In the diagram above we see some contour lines of the function, one of
which is marked '0'.  This is clearer using the following idiom:

```{r}
filled.contour(xvals,yvals,M>0,asp=1)
```

In the graph above, we see the positive region in one colour and the
negative region in another; where these meet is the solution to
$f(x,y)=0$.
    
## multivariate optimization

To minimize a multivariate function we need to use `optim()`,
which requires us to provide an initial value.  From the contour graph
we can see that there is a local minimum near $(0.1,-2)$ so we can use
that:

```{r}
optim(c(0.1,-2),f)
```

So the idiom above has found a minimum at $(0.32,-2.12)$.  To verify
this, we plot the function in a small region around this point:


```{r}
xvals <- seq(from = 0.32-0.1, to = 0.32+0.1,len=100)
yvals <- seq(from =-2.12-0.1, to =-2.12+0.1,len=100)
jj <- expand.grid(xvals,yvals)
M <- apply(jj,1,f)     # apply the function f() to jj
M <- matrix(M,100,100) # wrap M up into a matrix
par(pty="s")
contour(xvals,yvals,M,asp=1,xlim=range(xvals),ylim=range(yvals))
points(0.32,-2.12,pch=16,col='red')
```

In the above, the point marked is exactly in the center of the contour
lines, as it should be.  The function has another local minimum: can
you find it?
