---
output:
  pdf_document: default
  html_document: default
---
# Univariate integration {#univariate_integration}


Look at this figure:

\includegraphics{calculus2}

this shows the basic idea of integral calculus.  If $A(x)$ is the area
under the curve (that is, the area between the x-axis and the line
$y=f(x)$), up to point $x$, then we can see that


\[
\frac{dA(x)}{dx}=f(x)
\]


This is the fundamental theorem of calculus.  Again, this is an
informal treatment lacking in full rigour.  The formula amounts to
finding a function $A(x)$ which, when differentiated, is equal to our
original $f(x)$, which is known.  We usually write

\[
A(x) = \int f(x)dx
\]

and some examples follow.

* $f(x) = x^6\longrightarrow A(x)=\frac{1}{7}x^7$
* $f(x) = \sin(x)\longrightarrow A(x) = -\cos(x)$

Other examples are given in textbooks, and you may see a constant
added.  Sometimes we can use algebraic methods to calculate areas and
sometimes numerical methods are better. 

## Integration example

We will use a simple example, integrating $f(x)=x^2$ from $x=1$ to
$x=3$, and compare methods.

```{r}
x <- seq(from=0,to=3,len=100)
f <- function(x){x^2}
plot(x,f(x),type='l')
z <- seq(from=1,to=3,len=100)
jj <- c(x,rev(x))
polygon(c(z,rev(z)),c(z*0,f(rev(z))),col="red")
```

### Method 1: algebraic.

The integral of $x^2$ is $\frac{1}{3}x^3$ so the area under the curve
is $\left. \frac{1}{3}x^3\right|_{x=1}^{x=3}=9-1/3=8\frac{2}{3}$.
This is an exact value.

### Method 2: basic numerics

We use R, basically to evaluate $\sum f(x_i)dx$:

```{r}
dx <- 1e-3
x <- seq(from=1,to=3,by=dx)
sum(f(x)*dx)
```

showing a small discrepancy from the true value of $8.667$.  It is
possible to reduce the error by using a smaller value for ```dx```.

### Method 3, numerical integration

R has lots of very very sophisticated routines to integrate areas
using very accurate techniques:

```{r}
integrate(f,1,3)
```

showing a very accurate result.

