---
output:
  pdf_document: default
  html_document: default
---
# Multivariate integration {#multivariate_integration}



Suppose we have a function of two variables, $x$ and $y$ and wish to
find the volume between the function's surface and the plane $z=0$.
Consider again the function discussed in chapter 2:

\[
f(x,y) = x^2+y^2 + x + 4y + \sin(xy)
\]


```{r}
f <- function(v){
  x <- v[1]
  y <- v[2]
  return(x^2+y^2 + x + 4*y + sin(x*y))
}
```

We will visualise $f()$ using ```persp()```:

```{r}
xvals <- seq(from = -2, to = +2, len=100)
yvals <- seq(from = -2, to = +2, len=100)
jj <- expand.grid(xvals,yvals)

M <- apply(jj,1,f)     # apply the function f() to jj
M <- matrix(M,100,100) # wrap M up into a matrix 
persp(xvals,yvals,M)
```

To find the volume $V$ enclosed we need an iterated integral:


\[
I=
\int_{-2\leq x,y\leq 2}
\left(x^2+y^2 + x + 4y + \sin(xy)\right)dy\,dx
\]

[See Hubbard and Hubbard for a rigorous approach; the issue is not
straightforward].

Analytically 


$$
\begin{aligned}
I &=&
\int_{x=-2}^2
\int_{y=-2}^2
\left(x^2+y^2 + x + 4y + \sin(xy)\right)\,dy\,dx\\
 &=& \int_{x=-2}^2 \left(4x^2+4x+\frac{16}{3}\right)\,dx\\
 &=& \left[\frac{4}{3}x^3 +\frac{64}{3}\right]_{-2}^{+2}\\
 &=& \frac{128}{3}
\end{aligned}
$$

or $42\frac{2}{3}\simeq 42.67$.  We can accomplish this calculation
numerically using the ```adaptIntegrate()``` function of the
```cubature``` R package:

```{r}
library("cubature")
adaptIntegrate(f, c(-2,-2),c(2,2))
```

showing close agreement.
