# Linear algebra: matrices {#matrices}
 

A *matrix* is a block of numbers, as in

$$
\left(\begin{array}{ccc} 
0 & 5 & 75\\
3 & -1 & -3
\end{array}\right)
$$

We can view a matrix as a bunch of column vectors, written
side-by-side.  A matrix can have any number of rows or columns.
Matrices are *astonishingly* useful objects in mathematics and
statistics, and their study is well worth while.
You will not be surprised to learn that R includes a vast range of
functionality for working with matrices.

To create the matrix above in R, we could use:

```{r}
matrix(c(0,3,5,-1,75,3),2,3)
matrix(c(0,5,75,3,-1,-3),2,3,byrow=TRUE)
cbind(c(0,3),c(5,-1),c(75,-3))
rbind(c(0,5,75,3),c(-1,-3),2,3)
```

and indeed there are many other ways (arguments ```2``` and ```3```
refer to the number of rows and columns respectively).

## Index notation for matrices

Given a matrix $M$ as in 

$$
M=\left(\begin{array}{ccc} 
0 & 5 & 75\\
2 & 35 & 76\\
5 & -52 & 79\\
3 & -1 & -3
\end{array}\right)
$$

We can refer to individual elements using $M_{ij}$ which refers to row
$i$, column $j$.  Thus in the example above we would have $M_{11}=0$,
$M_{22}=35$, $M_{43}=-3$, and so on.  In R we say ```M[i,j]``` instead
of $M_{ij}$:

```{r}
M <- matrix(c(
  0 ,   5 , 75,
  2 ,  35 , 76,
  5 , -52 , 79,
  3 ,  -1 , -3),
  4,3,byrow=TRUE)
M
M[2,3]
M[1,1] <- 1000
M
```

## Matrix transpose

To take the transpose of a matrix, we exchange rows for columns.  We
write $A'$ or $A^T$ or sometimes $\operatorname{tr}(A)$ for the
transpose of $A$, and in R we use function ```t()```:

```{r}
M
t(M)
```

In index notation, if $B=A^T$ then $B_{ij}=A_{ji}$.

## Types of matrices

### The zero matrix

The zero matrix has all entries zero:

```{r}
matrix(0,5,6)
```

### Square matrices

A *square* matrix has the same number of rows as columns, as in:

$$
\left(\begin{array}{ccc} 
3 & 1  & 1\\ 
3 & 1  & 1\\ 
0 & 5 & 75\\
\end{array}\right)
$$

A *diagonal* matrix has nonzero entries only along the main diagonal
as in:


$$
\left(\begin{array}{cccc} 
3 &  0  & 0  & 0\\ 
0 & -1  & 0  & 0\\ 
0 &  0  & 7  & 0\\ 
0 &  0  & 0  & 2\\ 
\end{array}\right)
$$


In R:

```{r}
diag(c(3,-1,7,2))
```

If we have a very large matrix we indicate continuation using dots, as in


$$
\left(\begin{array}{cccc} 
1 &  0  & \cdots & 0\\ 
0 &  2  & \cdots  & 0\\ 
\vdots &\vdots &\ddots&\vdots\\ 
0 &  0 &\cdots&8
\end{array}\right)
$$

In R:

```{r}
diag(1:8)
```

A diagonal matrix must be square.

### Symmetrical matrices

A symmetrical matrix is equal to its own transpose: matrix $A$ is
symmetrical if $A^T=A$.  A symmetrical matrix must be square.  For
example

```{r}
A <-    # define a symmetrical matrix...
matrix(c(
  1,4,5,
  4,0,2,
  5,2,5),3,3)
A
all(A ==  t(A)) #... and verify that it is symmetrical
```


We will see other types of matrices later in the chapter.


## Vector space operations on matrices

### Scalar multiplication

Matrices multiplied by a scalar as one might
expect:

```{r}
M <- matrix(1:30,5,6)
M
M*7
```

### Matrix addition and subtraction

Matrices *of the same size*  may be added together:

```{r}
matrix(1:6,2,3) + matrix(100,2,3)
```

Adding matrices of different sizes gives an error:

```{r}
try(matrix(1:6,2,3) + matrix(0,2,4))
```


## Matrix multiplication

Matrices can be multiplied together.  The rules for matrix
multiplication seem strange, but make a powerful analyical method.
Given two matrices $A$ and $B$, we define the *matrix product* $AB$ as
follows.

If $A$ is $n\times m$ (that is, $A$ has $n$ rows and $m$ columns), and
$B$ is $m\times p$, then the matrix produce $C=AB$ is defined as
having elements

$$
c_{ij} = a_{i1}b_{1j}+a_{i2}b_{2j}+\cdots+a_{im}b_{mj}=
\sum_{k=1}^ma_{ik}b_{kj}
=\sum a_{ik}b_{kj}
$$

Note that the number of _columns_ of $A$ must be equal to the number
of _rows_ of $B$.  See the following diagram (taken from Wikipedia):

\includegraphics{626px-Matrix_multiplication_diagram_2.svg}

For example, in R we might say

```{r}
A <- matrix(1:6,2,3)
B <- matrix(15:1,3,5)
A
B
A %*% B
```

Note carefully that matrix multiplication uses ```%*%``` (an ordinary
```*``` means pointwise multiplication.  Never ever ever ever ever
ever ever ever ever ever ever ever ever ever ever ever ever ever ever
ever ever ever ever ever ever ever ever ever ever ever ever ever ever
ever ever ever ever ever ever ever ever ever ever ever ever ever ever
ever ever ever ever ever ever ever ever ever ever use this, unless you
have a Fields medal, and even then don't use it.  Pointwise
multiplication will poison your mind, give you a parking ticket,
borrow your car and not fill it up with petrol, tread on your shoes,
find your fridge and drink all your beer).

Note even more carefully that matrix multiplication is not commutative
in the sense that, in general, $AB\neq BA$.  In case above, $BA$ is
not even defined:

```{r}
try(B %*% A)
```

(in the above, the result is not defined because the matrices do not
satisfy the rule for matrix multiplication, viz ``number of _columns_
of the first matrix must be equal to the number of _rows_ of the
second''.

Even when $AB$ and $B$ are both defined (a common case is when $A$ and
$B$ are both square matrices of the same size), the result will
differ:


```{r}
A <- matrix(rnorm(25),5,5)
B <- matrix(rnorm(25),5,5)
A %*% B
B %*% A
```